tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-13 paper.tex
Minor edits from Gertjan + Daiki

\abstract{
\textbf{Summary:}
Ancestral recombination graphs (ARGs) encode the ensemble of correlated
genealogical trees arising from recombination in a compact and efficient
structure, and are of fundamental importance in population and statistical
genetics. Recent breakthroughs have made it possible to simulate and
infer ARGs at the population scale, and there is now intense interest
in applying ARG-based methods across a broad range of applications.
Genome-wide association studies (GWAS) are one of the most imporant
application areas, with ARG-based methods already providing substantial
improvements over the state-of-the-art. The ability to efficiently
simulate quantitative traits is critical to developing
and evaluating new GWAS methods, but existing simulators do not scale
to modern datasets and cannot be directly applied to ARGs.
We present the \texttt{tstrait} package,
an open-source Python library to simulate
quantitative traits on ARGs, and show how this user-friendly software
can be easily applied to vast, population-scale datasets.
\\\textbf{Availability and Implementation:} \texttt{tstrait} is available
for download on the Python Package Index.
Full documentation with examples and workflow
templates is available on \url{https://tskit.dev/tstrait/docs/}, and the
development version is maintained on Github
(\url{https://github.com/tskit-dev/tstrait}).\\ \textbf{Contact:}
\href{daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}\\
}

@@

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been identified~\citep{locke2015,ishigaki2022,
mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
Despite these successes and the ongoing collection of population-scale
Biobank data~\citep{tanjo2021practical}, there are substantial open
questions around GWAS methology.
In particular, correctly accounting for population stratification
remain challenging~\citep{uffelmann2021}.

Ancestral recombination graphs (ARGs) encode the interwoven paths
of genetic inheritance arising from recombination~\citep{wong2023general}, and recent
breakthroughs have opened many new possibilities for their
practical application~\citep{lewanski2023era}.
Building on these breakthroughs, substantial progress has been made
on statistical genetics questions.
For example,  ARG-based methods
can detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
can better detect causal loci in quantitative-trait locus mapping~\citep{link2023tree};
and can provide a sparse and efficient model of linkage disequilibirium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

@@

The first step of simulating data for evaluating GWAS methods,
generating realistic genetic variation datasets of millions
of individuals, is therefore at a high-level of sophistication
and more than sufficient to support ongoing research.
The second step, of generating the actual quantiative traits,
is currently lagging far behind.
Although there are multiple packages to simulate genotypic data
concurrently with quantitative traits~\citep{gaynor2021,haller2023},
and to simulate quantitative traits from a given set of genetic
sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
handle the millions of samples now required.
[TODO discuss bespoke stuff like Martin et al]
In particular, there is no software package available to simulate quantitative
traits directly from the efficient ARG representation used by
modern population genetic simulators.

@@

\subsection{Effect Sizes}
The first step of quantitative trait simulation in \texttt{tstrait} is to
define the trait model distribution, as follows:
\begin{lstlisting}[language=Python]
model = tstrait.trait_model(
    distribution='normal', mean=0, var=1)
\end{lstlisting}
\texttt{Tstrait} currently supports effect size simulation from five
commonly used univariate distributions~\citep{gaynor2021,haller2023}
and one multivariate distribution.
The multivariate normal distribution is used to simulate multiple
pleiotropic traits.
The trait model is used as an input of \texttt{sim\_trait} function to simulate
effect sizes, along with the input ARG and number of causal sites:
\begin{lstlisting}[language=Python]
trait_df = tstrait.sim_trait(
    arg, num_causal=5, model=model)
\end{lstlisting}
The output, \texttt{trait\_df} is a \texttt{pandas} dataframe
describing the simulated effect sizes and corresponding
causal site IDs.

>>>>

\abstract{
\textbf{Summary:}
Ancestral recombination graphs (ARGs) encode the ensemble of correlated
genealogical trees arising from recombination in a compact and efficient
structure, and are of fundamental importance in population and statistical
genetics. Recent breakthroughs have made it possible to simulate and
infer ARGs at the population scale, and there is now intense interest
in applying ARG-based methods across a broad range of applications.
Genome-wide association studies (GWAS) are one of the most important
application areas, with ARG-based methods already providing substantial
improvements over the state-of-the-art. The ability to efficiently
simulate quantitative traits is critical to developing
and evaluating new GWAS methods, but existing simulators do not scale
to modern datasets and cannot be directly applied to ARGs.
We present the \texttt{tstrait} package,
an open-source Python library to simulate
quantitative traits on ARGs, and show how this user-friendly software
can be easily applied to vast, population-scale datasets.
\\\textbf{Availability and Implementation:} \texttt{tstrait} is available
for download on the Python Package Index.
Full documentation with examples and workflow
templates is available on \url{https://tskit.dev/tstrait/docs/}, and the
development version is maintained on Github
(\url{https://github.com/tskit-dev/tstrait}).\\ \textbf{Contact:}
\href{daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}\\
}

@@

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been identified~\citep{locke2015,ishigaki2022,
mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
Despite these successes and the ongoing collection of population-scale
Biobank data~\citep{tanjo2021practical}, there are substantial open
questions around GWAS methology.
% TODO rephrase
In particular, correctly accounting for population stratification
remains challenging~\citep{uffelmann2021}.

Ancestral recombination graphs (ARGs) encode the interwoven paths
of genetic inheritance arising from recombination~\citep{wong2023general}, and recent
breakthroughs have opened many new possibilities for their
practical application~\citep{lewanski2023era}.
Building on these breakthroughs, substantial progress has been made
on statistical genetics questions.
For example,  ARG-based methods
can detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
have better power to detect causal loci in quantitative-trait locus mapping~\citep{link2023tree};
and can provide a sparse and efficient model of linkage disequilibirium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

@@

The first step of simulating data for evaluating GWAS methods,
generating realistic genetic variation datasets of millions
of individuals, is therefore at a high-level of sophistication
and more than sufficient to support ongoing research.
The second step, of generating the actual quantiative traits,
is currently lagging far behind.
Although there are multiple packages to simulate genotypic data
concurrently with quantitative traits~\citep{gaynor2021,haller2023},
and to simulate quantitative traits from a given set of genetic
sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
handle the millions of individuals now required.
[TODO discuss bespoke stuff like Martin et al]
In particular, there is no software package available to simulate quantitative
traits directly from the efficient ARG representation used by
modern population genetic simulators.

@@

\subsection{Effect Sizes}
The first step of a quantitative trait simulation in \texttt{tstrait} is to
define the trait model distribution, as follows:
\begin{lstlisting}[language=Python]
model = tstrait.trait_model(
    distribution='normal', mean=0, var=1)
\end{lstlisting}
\texttt{Tstrait} currently supports effect size simulation from five
commonly used univariate distributions~\citep{gaynor2021,haller2023}
and one multivariate distribution.
The multivariate normal distribution is used to simulate multiple
pleiotropic traits.
The trait model is used as an input of \texttt{sim\_trait} function to simulate
effect sizes, along with the input ARG and number of causal sites:
\begin{lstlisting}[language=Python]
trait_df = tstrait.sim_trait(
    arg, num_causal=5, model=model)
\end{lstlisting}
The output, \texttt{trait\_df} is a \texttt{pandas} dataframe
describing the simulated effect sizes and corresponding
causal site IDs.
